A Recipe for Scavenging - the natural history of a behaviour

### 2016-03-22 natural history of scavenging.tex
new frame for the paper
aerial and terrestrial sections changed to reflect this

\title{A recipe for scavenging and natural history}
\running{Scavenging in vertebrates}

@@

\subsection*{The Difficulty of Scavenging} %KH: Make that the template for all the next sections
The chief hurdle to scavenging is finding a sufficient quantity of food, the occurence of which is difficult to predict in space and time.
Once found, the scavenger has to then overcome the agents of decay produced by the action of microorganisms on the carcass \citep{ruxton2014fruit}.
The idea of scrounging from predator kills is undermined by studies showing that in the  majority of ecosystems more animals die from disease and starvation than predation \citep{benbow2015introduction}.
Thus, any animal existing as a scavenger must maximise its detection capabilities and minimise its locomotory costs \citep{ruxton2004obligate}.
The habitat must also be productive enough to sustain an animal biomass that will eventually produce carcasses.

@@

\section*{Aerial Scavengers}
Scavenging should be particularly attractive to avian predators compared to mammals.
Solitary mammalian predators can kill prey up to the same body mass as themselves and sometimes an order of magnitude heavier (e.g. socially hunting lions \citep{owen2008predator}).
In contrast, birds of prey tend to kill prey smaller than themselves \citep{slagsvold2007prey}.
This is likely due to their need to kill prey that they can fly away with, as well as the risk of injury being higher (which carries a higher mortality risk) for a bird than a mammal.
Thus for birds, scavenging means they can exploit species that would otherwise be too big for them to kill.

Vultures represent the best known scavengers on Earth.
These birds consist of two convergent groups, from the old and the new world and represent the only example of obligate vertebrate scavengers today.
The families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene (66 - 56 Mya) \citep{Jetz2012, Jarvis2014}.
Given their unique position, they have been extensively studied to determine what adaptations they possess that allows them to so flourish in this niche.
Underscoring their prowess are studies such as \cite{ogada2012effects} who note a three fold increase in carcass persistence times in the absence of vultures.
As such, we can begin by exploring the adapdations and the environments of vultures to draw comparisons with other scavenging species and \textit{their} environments.

@@

Species capable of flight have effectively added an extra spatial dimension, i.e. the vertical component, to their sensory environment over land animals.
This allows them to look down on a landscape where they are unencumbered by obstacles that would obstruct the view of a terrestrial scavenger.
Such an ability has obvious benefits in detecting carrion.
Vultures are known to have impressive visual acuity with one estimate indicating Lappet-faced Vultures (\textit{Torgos tracheliotus}) are capable of detecting a 2 metre carcass over 10 km away \citep{spiegel2013factors}.
We know that many birds exist as facultative scavengers; storks, eagles, corvids, are all known to take substantial quantities of carrion in their diet.
And eagles in particular are known to have highly developed visual abilities.
It follows from this that the evolution of flight allowed aerial animals to detect far more carrion than their terrestrial counterparts \citep{AR:AR22815}.

Aside from sight, three species within the new world family Cathartidae, (genus \textit{Cathartes}), have well developed olfactory systems \citep{AR:AR22815}.
Among them are the Turkey Vultures (\textit{Cathartes aura}) which were able to locate 90\% of baits set out in a tropical forest \citep{houston1986olfaction}.
This would be impossible for the visually reliant old world species.
This point illustrates how the environment can impact search efficiency depending on the sensory system that's used.

Flight is also a cheaper means of locomotion than running \citep{tucker1975energetic}.
This advantage can be extended further in larger species by engaging in soaring instead of flapping flight, which is even cheaper still (approximately twice BMR) \citep{hedenstrom1993migration}.
The benefits this confers are clear from the information we have on the enormous foraging ranges of many vultures \citep{spiegel2013factors}.
Clearly, it would be pointless to have incredible detection abilites and not have a cost efficient movement to benefit from it.
Although, as with detection ability, the environment has a role to play here.
Vultures and eagles tend to soar using thermals and if these air pockets don't form, say on a cloudy day, the bird is grounded \citep{mundy1992vultures}.
In many habitats (e.g. the arctic) it is simply not possible for sufficiently powerful thermals to form and as a consequence large-bodied vultures cannot exist.
The upshot of this is that terrestrial carnivores like bears and wolves take more carrion \citep{devault2003scavenging}.

Avian flight originates in the Late Jurassic (163.5-145 Million years ago - Mya), conincident with the fossils of \textit{Archaeopteryx lithographica} so many of these benefits would have been realised from that point on for carnivorous birds.
However, vertebrate flight is much older than this where pterosaurs predate bird origins by a considerable margin in the Late Triassic (235-201.3 Mya).
Scavenging in this diverse group has been hypothesied many times before.
Certain clades of these animals could reach enormous sizes (e.g. Azhdarchids with wingspans of 11 metres \citep{witton2010size}) and look to have engaged in soaring flight \citep{witton2010size}.
Although \cite{witton2008reappraisal} argued that neck inflexibility and straight, rather than hooked jaw morphology points against their existing as \textit{obligate} scavengers, Azhdarchid terrestrial proficency indicates they would have been comfortable foraging on the ground.
Indeed, extant Marabou Storks (\textit{Leptoptilos crumenifer}) %TG: Check if that's the one
have a comparable morphology and are noted facultative scavengers so it is reasonable to believe that certain pterosaurs behaved similarly.

Large body size confers substantial dominance and starvation resistance benefits \citep{ruxton2004obligate}.
Thus, we would expect scavengers to have this trait selected for even in the case of weight-constrained fliers.
Cinereous Vultures (\textit{Aegypius monachus}) and condors (\textit{Vultur gryphus}, \textit{Gymnogyps californianus}) all have body masses that can exceed 10 kg and represent some of the heaviest bird species capable of flight \citep{ferguson2001raptors,donazar2002effects}.
And as we have noted the Azhdarchid pterosaurs were far bigger again, with estimated body masses of over 200 kg \citep{witton2010size}.

The only other vertebrate group capable of powered flight are the bats where scavenging has not been recorded to our knowledge.
Their visual acuity is famously poor and echolocation would not lend itself to discovering immobile carrion.
Their small size and poor terrestrial ability would also count against them at a carcass \citep{riskin2006terrestrial}.
The bat fossil record is notoriously poor owing to their fragile skeletons so we are unable to determine if extinct species were more suited to this lifestyle \citep{eiting2009global}.
%TG: Two points here: one speculation for the non-scavenging behaviour could be because of diet.
Big bats (which are better suited for scavenging, following previous paragraph) are essentially frugivores and therefore probably lack of adaptations for digesting the meat.
Also carnivores bats are mainly the microbats that are insectivores.
Maybe they scavenge on dead insects but here we're talking about scavenging on real meat right?
%TG: Second point: bats fossil record is of course not so good but it's actually way better than birds! Most of the data we have is on teeth of course but it's probably enough for having a rough body mass and diet estimate! And that's much better than birds where you're stuck with just some fragments of long bones.

\section*{Terrestrial Scavengers}
A simplification of terrestrial, vertebrate scavengers is one of them existing in a two-dimensional plane while foraging for carrion directly.
They can detect carcasses at a range that is defined by the radius of their sensory organs, usually the visual and olfactory senses.
As a consequence, they have a much more restricted view of the landscape than do aerial foragers.
No contemporary terrestrial vertebrate exists as an obligate scavenger but most if not all are facultative scavengers to some extent \citep{beasley2015vertebrates}.
\cite{ruxton2004obligate} offer a reason for this in that the traits that allow for vultures to exist as scavengers undermined their ability to hunt but that the same forces have not prevented mammals from doing so.
Interestingly, the same authors calculated that ``a 1 tonne mammal or reptile, in an ecosystem yielding carrion at densities similar to the current Serengeti, could have met its energy requirements if it could detect carrion over a distance of the order of 400–500 m."\citep{ruxton2004obligate}.

Terrestrial scavenging in the mammals is probably best known in an African context where hyenas, jackals and lions all take sizable proportions of carrion in their diet.
In the spotted hyena (\textit{Crocuta crocuta}), striped hyena (\textit{Hyaena hyaena}) and brown hyena (\textit{Hyaena brunnea}) it can be over 90\% \citep{jones2015african}.
Therefore, we can again use these species as our efficient terrestrial scavengers to compare with other forms.

Similar to vultures they have well developed sensory organs, whereby they can smell a rotting carcass 4 km away and can hear the vocalisations of conspecifics at a distance of 10 km \citep{mills1989comparative}.
They have a characteristic ``rocking horse gait"  which allows them to cover great distances efficiently \citep{jones2015african}.
A major difficulty for terrestrial scavengers is competition with vultures.
Noctural behaviour in the Hyaenidae in general has been put forth as an adaptation to reduce competition with these exclusively diurnal birds \citep{gittleman2013carnivore}.
The bone crushing ability of hyenas reveals another useful scavenger trait.
Since carrion is not dispatched directly, often the most easily accessible and choicest components of the carcass will be missing or, if present, will be fought over.
Being able to extract nutrients from remnants gives the scavenger a great advantage.
Osteophagy is known across a range of terrestrial carnivores and given some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, it makes skeletal remains an enticing resource \citep{brown1989study}.
This ability reached its zenith among hyenas with the evolution of the 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep{palmqvist2011giant}.
Some work on extinct sabretooths suggests they may have left a large amount of food for would-be scavengers because of their unique skull morphology.
As a result, the decline of Machairodontinae sabretooths has been offered as an explanation for the extinction of \textit{P brevirostris} \citep{palmqvist2011giant}.

The order Carnivora sees its origins in the Middle Eocene (56-33.9 Mya) where it split into the Caniforma and Feliforma.
And many of the aforesaid adapations for scavenging are found in these other major terrestrial mammalian carnviores.
Though the specific mix of features realised in hyenas suggest this is the model organism for terrestrial scavenging among mammals in the past.
For example, the bone-crushing dogs that evolved during the Oligocene (subfamily Borophaginae) have been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.
Certainly, it is profitable to compare extant terrestrial species to their prehistoric forebears given the dominance of mammalian carnviores since the Eocene.

@@

By contrast to mammals, a successful reptilian scavenger requires a far different set of adapations.
Modern forms are ectothermic, limiting their activity periods.
This is exacerbated by the sprawling gait seen in many lizards which results in Carrier's Constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}.
This manifests itself in aspects such as maximum sustainable speed where an equivalent mammal has a six to seven fold increase \citep{ruben1995evolution}.
A lower metabolism does give reptiles an advantage however, in that over the course of a year their food requirements can be 30 times smaller than an endotherm of equal size \citep{Nagy1621}.
Any adaptations that reduce energetic costs are likely to be selected in scavengers.
\cite{devault2002scavenging} suggest this is an avenue for scavenging in snakes because they ``exhibit  exceedingly  low  maintenance  metabolisms,  and most  can  survive  on  a  few  scant  feedings per year.
It  is, therefore, possible for snakes to rely largely  on  infrequent,  less  energy-rich  meals." In the same review the authors found occurrences of scavenging spread across five families of snakes and stated that this behaviour is ``far more common than currently acknowledged."\citep{devault2002scavenging}.
The same reasoning can be applied to crocodlies and their allies \citep{forrest2003evidence}.

Unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous, theropod dinosaurs.
These animals ranged from the chicken-sized to the whale-sized all of which were bipedal.
They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}.
Of relevance, are the questions that still persist about their metabolism \citep{grady2014evidence} and sensory perception \citep{farlow1994speculations}.
We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{weishampel2004dinosauria,depalma2013physical}.
Much work has focused on the existence of the behaviour in \textit{Tyrannosaurus rex} \citep{ruxton2003could,carbone2011intra} but a recent energetics study investigated the likely prevelance of scavenging across a range of body sizes.
In it the authors demonstrated that species of intermediate body masses (approximatively 500 kg) would have gained the most benefit from scavenging.
This was the result of gut capacity limitations and the effects of competition at the carcass.
At the larger extreme this owes to the fact that gut capacity doesn't scale isometrically with body mass so the benefits of greater mass level off; there's only so much food an individual can consume at a single sitting \citep{calder1996size}.
For the smaller species, larger competitors would have prevented their access to carrion.

As we discussed for the case of Cenozoic carnivores, osteophagy could be extremely beneficial to a scavenger.

In Mesozoic systems some extremely large theropod dinosaurs had a morphology which suggests an ability to process bone e.g. the robust skull and dentition of \textit{T. rex} \citep{hone2010feeding}.
There is direct evidence that \textit{T. rex} did this in the form of distinctive wear marks on its tooth apices \citep{farlow1994wear,schubert2005wear} and the presence of bone fragments in its coprolites \citep{chin1998king}.
The animal also had an enormous bite force, with one estimate putting it at 57000 Newtons \citep{bates2012estimating}.
This is noted as being powerful enough to break open skeletal material \citep{rayfield2001cranial}.
Osteophagy may have been even more viable during this era because the body mass distribution of herbviores tended to be skewed towards larger sizes \citep{10.1371/journal.pone.0051925}.
When we couple this with the fact that skeletal mass scales greater than linearly with body mass \citep{prange1979scaling} there would have been a lot of bone material to consume in the environment provided an animal had the biology to process it \citep{chure1997one}.

Of course, tetrapod terrestrial dominance predates the evolution of the dinosaurs.
It is in the Permian, almost 300 millions years ago, that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}.
The absence of flying vertebrates in the Palaeozoic may have permitted terrestrial forms to take in a higher proportion of carrion in their diet.
As noted earlier, a system similar to the Serengeti in productivity could have supported a terrestrial scavenger \citep{ruxton2004obligate}.
Indeed, in systems that were dominated by large ectothermic or mesothermic vertebrates 
the same primary productivity would have supported a greater biomass \citep{mcnab2009resources}.
The upshot of this is there was a higher biomass of herbivores dying and offering scavenging opportunities.
Predators were large bodied too compared to extant mammalian predators \citep{mcnab2009resources}, and so, especially if they were ectothermic, could last longer between meals rendering scavenging a more attractive behaviour relative to predation.

%TG: To be modified and inserted somewhere:
Scavenging behaviour might have evolved as soon as the first terrestrial tetrapods evolved.
In fact, some of the earlier tetrapods tracks dating back to the early Middle Devonian (393.3 - 387.7 Mya) where found in intertidal environments \citep{Niedzwiedzki2009}.
These environments are isolated from marine systems twice a day leaving potential carrions unexploited by marine vertebrates.
\cite{Niedzwiedzki2009} suggests that these environments ``would thus have allowed marine ancestors of tetrapods gradually to acquire terrestrial competence while accessing a new and essentially untouched resource.''
%TG: orginal sentence "The intertidal environment provides a ready food source of stranded marine animals on a twice-daily basis, in the immediate vicinity of the sea, and would thus have allowed marine ancestors of tetrapods gradually to acquire terrestrial competence while accessing a new and essentially untouched resource."

>>>>

\title{A Recipe for Scavenging - the natural history of a behaviour}
\running{Scavenging in vertebrates}

@@

\subsection*{The Difficulty of Scavenging} %KH: Make that the template for all the next sections
The chief hurdle to scavenging is finding a sufficient quantity of food, the occurence of which is difficult to predict in space and time. Thus, any animal existing as a scavenger must minimise its locomotory costs and maximise its detection capabilities \citep{ruxton2004obligate}.
Once found, the scavenger has to process the carrion and overcome the agents of decay produced by the action of microorganisms on the carcass \citep{ruxton2014fruit}. 
The habitat must also be productive enough to sustain an animal biomass that will eventually produce carcasses.
We can draw on the image of a scavenger moving through its environment, searching for food and trying to process it efficiently as we explore the prevalence of this behaviour through time. 
% The idea of scrounging from predator kills is undermined by studies showing that in the  majority of ecosystems more animals die from disease and starvation than predation \citep{benbow2015introduction}.

@@

\section*{Aerial Scavengers}
Vultures represent the best known scavengers on Earth.
These birds consist of two convergent groups, from the old and the new world and represent the only example of obligate vertebrate scavengers today.
The families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene (66 - 56 Mya) \citep{Jetz2012, Jarvis2014}.
Given their unique position, they have been extensively studied to determine what adaptations they possess that allows them to so flourish in this niche.
Underscoring their prowess are studies such as \cite{ogada2012effects} who note a three fold increase in carcass persistence times in the absence of vultures.
As such, we can begin by exploring the adapdations and the environments of vultures to draw comparisons with other scavenging species and \textit{their} environments.

\subsection*{Locomotion}
Flight is a cheaper means of locomotion than running \citep{tucker1975energetic}.
This advantage can be extended further in larger species by engaging in soaring instead of flapping flight, which is even cheaper still (approximately twice BMR) \citep{hedenstrom1993migration}.
The benefits this confers are clear from the information we have on the enormous foraging ranges of many vultures \citep{spiegel2013factors}.
Avian flight originates in the Late Jurassic (163.5-145 Million years ago - Mya), conincident with the fossils of \textit{Archaeopteryx lithographica} so many of these benefits would have been realised from that point on for carnivorous birds.
However, vertebrate flight is much older than this where pterosaurs predate bird origins by a considerable margin in the Late Triassic (235-201.3 Mya).
Scavenging in this diverse group has been hypothesied many times before.
Certain clades of these animals could reach enormous sizes (e.g. Azhdarchids with wingspans of 11 metres \citep{witton2010size}) and look to have engaged in soaring flight \citep{witton2010size}.

The only other vertebrate group capable of powered flight are the bats where scavenging has not been recorded to our knowledge.

\subsection*{Detection}
Species capable of flight have effectively added an extra spatial dimension, i.e. the vertical component, to their sensory environment over land animals.
This allows them to look down on a landscape where they are unencumbered by obstacles that would obstruct the view of a terrestrial scavenger.
Such an ability has obvious benefits in detecting carrion.
Vultures are known to have impressive visual acuity with one estimate indicating Lappet-faced Vultures (\textit{Torgos tracheliotus}) are capable of detecting a 2 metre carcass over 10 km away \citep{spiegel2013factors}.
We know that many birds exist as facultative scavengers; storks, eagles, corvids, are all known to take substantial quantities of carrion in their diet.
And eagles in particular are known to have highly developed visual abilities.
It follows from this that the evolution of flight allowed aerial animals to detect far more carrion than their terrestrial counterparts \citep{AR:AR22815}.

Aside from sight, three species within the new world family Cathartidae, (genus \textit{Cathartes}), have well developed olfactory systems \citep{AR:AR22815}.
Among them are the Turkey Vultures (\textit{Cathartes aura}) which were able to locate 90\% of baits set out in a tropical forest \citep{houston1986olfaction}.
% This would be impossible for the visually reliant old world species.

Clearly, it would be pointless to have incredible ranging abilites and not have the sensory architecture to benefit from it.

\subsection*{Processing}
Scavenging should be particularly attractive to avian predators compared to mammals.
Solitary mammalian predators can kill prey up to the same body mass as themselves and sometimes an order of magnitude heavier (e.g. socially hunting lions \citep{owen2008predator}).
In contrast, birds of prey tend to kill prey smaller than themselves \citep{slagsvold2007prey}.
This is likely due to their need to kill prey that they can fly away with, as well as the risk of injury being higher (which carries a higher mortality risk) for a bird than a mammal.
Thus for birds, scavenging means they can exploit species that would otherwise be too big for them to kill.

Large body size confers substantial dominance and starvation-resistance benefits \citep{ruxton2004obligate}.
Thus, we would expect scavengers to have this trait selected for even in the case of weight-constrained fliers.
Cinereous Vultures (\textit{Aegypius monachus}) and condors (\textit{Vultur gryphus}, \textit{Gymnogyps californianus}) all have body masses that can exceed 10 kg and represent some of the heaviest bird species capable of flight \citep{ferguson2001raptors,donazar2002effects}.
And as we have noted the Azhdarchid pterosaurs were far bigger again, with estimated body masses of over 200 kg \citep{witton2010size}.

Although \cite{witton2008reappraisal} argued that neck inflexibility and straight, rather than hooked jaw morphology points against pterosaurs existing as \textit{obligate} scavengers, Azhdarchid terrestrial proficency indicates they would have been comfortable foraging on the ground.
Indeed, extant Marabou Storks (\textit{Leptoptilos crumenifer}) %TG: Check if that's the one
have a comparable morphology and are noted facultative scavengers so it is reasonable to believe that certain pterosaurs behaved similarly.

Returning to bats, their visual acuity is famously poor and echolocation would not lend itself to discovering immobile carrion.
Their small size and poor terrestrial ability would also count against them at a carcass \citep{riskin2006terrestrial}.
The bat fossil record is notoriously poor owing to their fragile skeletons so we are unable to determine if extinct species were more suited to this lifestyle \citep{eiting2009global}.
%TG: Two points here: one speculation for the non-scavenging behaviour could be because of diet.
Big bats (which are better suited for scavenging, following our previous argument) are typically frugivores and therefore probably lack of adaptations for digesting the meat.
Also carnivorous bats are mainly the microbats which are insectivorous.
% Maybe they scavenge on dead insects but here we're talking about scavenging on real meat right?
%TG: Second point: bats fossil record is of course not so good but it's actually way better than birds! Most of the data we have is on teeth of course but it's probably enough for having a rough body mass and diet estimate! And that's much better than birds where you're stuck with just some fragments of long bones.

\subsection*{Environment}
Vultures and eagles tend to soar using thermals and if these air pockets don't form, say on a cloudy day, the bird is grounded \citep{mundy1992vultures}.
In many habitats (e.g. the arctic) it is simply not possible for sufficiently powerful thermals to form and as a consequence large-bodied vultures cannot exist.
The upshot of this is that terrestrial carnivores like bears and wolves take more carrion \citep{devault2003scavenging}.
The use of different sensory systems also illustrates the impact of the environment. 
The relatively open savanna systems of Africa are well suited to a visually dependent vulture whereas more forested areas would select for species that have a well developed oflfactory system  \citep{houston1986olfaction}.

% Although, as with detection ability, the environment has a role to play here.

@@

% This point illustrates how the environment can impact search efficiency depending on the sensory system that's used.

\section*{Terrestrial Scavengers}
Terrestrial scavenging in the mammals is probably best known in an African context where hyenas, jackals and lions all take sizable proportions of carrion in their diet.
In the spotted hyena (\textit{Crocuta crocuta}), striped hyena (\textit{Hyaena hyaena}) and brown hyena (\textit{Hyaena brunnea}) it can be over 90\% \citep{jones2015african}.
Therefore, we can again use these species as our efficient terrestrial scavengers to compare with other forms.

No contemporary terrestrial vertebrate exists as an obligate scavenger but most if not all are facultative scavengers to some extent \citep{beasley2015vertebrates}.
\cite{ruxton2004obligate} offer a reason for this in that the traits that allow for vultures to exist as scavengers undermined their ability to hunt but that the same forces have not prevented mammals from doing so.
Interestingly, the same authors calculated that ``a 1 tonne mammal or reptile, in an ecosystem yielding carrion at densities similar to the current Serengeti, could have met its energy requirements if it could detect carrion over a distance of the order of 400–500 m."\citep{ruxton2004obligate}.

\subsection*{Locomotion}
As endotherms, mammals can sustain long bouts of energetically expensive activity. 
By contrast, modern reptiles are ectothermic, limiting their activity periods.
% a successful reptilian scavenger requires a far different set of adapations.
This is exacerbated by the sprawling gait seen in many lizards which results in Carrier's Constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}.
This manifests itself in aspects such as maximum sustainable speed where an equivalent mammal has a six to seven fold increase \citep{ruben1995evolution}.
To quantify this effect with a simple example we can turn to some allometric relationships relating sustainable travelling speed to body mass.
In the case of mammals and reptiles these are 1.15 * body mass (kg) \textsuperscript{0.12} and 0.23 * body mass (kg) \textsuperscript{0.12}
% respectively \citep{ruxton2004obligate}.
We can insert these into a foraging radius model ((duration * speed)/2)/1000 for a 12 hour foraging day which shows that while a 10 kg reptile can range 6.5 km an equally sized mammal can range nearly 33 km \citep{Enstipp2006Energetics}.
Hyenas extend this advantage with their characteristic ``rocking horse gait"  which allows them to cover great distances efficiently \citep{jones2015african}.

Unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous, theropod dinosaurs.
These animals ranged from the chicken-sized to the whale-sized all of which were bipedal.
They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}.
Of relevance, are the questions that still persist about their metabolism \citep{grady2014evidence}. 
We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{weishampel2004dinosauria,depalma2013physical}.

Of course, tetrapod terrestrial dominance predates the evolution of the dinosaurs.
It is in the Permian, almost 300 millions years ago, that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}.

\subsection*{Detection}
A simplification of terrestrial, vertebrate scavengers in sensory terms is one of them existing in a two-dimensional plane while foraging for carrion directly.
They can detect carcasses at a range that is defined by the radius of their sensory organs, usually the visual and olfactory senses.
As a consequence, they have a much more restricted view of the landscape than do aerial foragers.
Similar to vultures hyenas have well developed sensory organs, whereby they can smell a rotting carcass 4 km away and can hear the vocalisations of conspecifics at a distance of 10 km \citep{mills1989comparative}. 
Using the same approach \cite{spiegel2013factors} used to determine the visual acuity of vultures we found a Spotted hyena could resolve a 2 metre target at 1 km distance.

\subsection*{Proceesing}
The bone crushing ability of hyenas reveals another useful scavenger trait.
Since carrion is not dispatched directly, often the most easily accessible and choicest components of the carcass will be missing or, if present, will be fought over.
Being able to extract nutrients from remnants gives the scavenger a great advantage.
Osteophagy is known across a range of terrestrial carnivores and given some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, it makes skeletal remains an enticing resource \citep{brown1989study}.
This ability reached its zenith among hyenas with the evolution of the 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep{palmqvist2011giant}.
Some work on extinct sabretooths suggests they may have left a large amount of food for would-be scavengers because of their unique skull morphology.
As a result, the decline of Machairodontinae sabretooths has been offered as an explanation for the extinction of \textit{P brevirostris} \citep{palmqvist2011giant}.
The order Carnivora sees its origins in the Middle Eocene (56-33.9 Mya) where it split into the Caniforma and Feliforma.
And many of the aforesaid adapations for scavenging are found in these other major terrestrial mammalian carnviores.
Though the specific mix of features realised in hyenas suggest this is the model organism for terrestrial scavenging among mammals in the past.
For example, the bone-crushing dogs that evolved during the Oligocene (subfamily Borophaginae) have been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.
Certainly, it is profitable to compare extant terrestrial species to their prehistoric forebears given the dominance of mammalian carnviores since the Eocene.

@@

When it comes to reptiles they possees an advantage here, in that over the course of a year their food requirements can be 30 times smaller than an endotherm of equal size \citep{Nagy1621}.
Any adaptations that reduce energetic costs are likely to be selected in scavengers.
\cite{devault2002scavenging} suggest this is an avenue for scavenging in snakes because they ``exhibit  exceedingly  low  maintenance  metabolisms,  and most  can  survive  on  a  few  scant  feedings per year.
It  is, therefore, possible for snakes to rely largely  on  infrequent,  less  energy-rich  meals." In the same review the authors found occurrences of scavenging spread across five families of snakes and stated that this behaviour is ``far more common than currently acknowledged."\citep{devault2002scavenging}.
The same reasoning can be applied to crocodlies and their allies \citep{forrest2003evidence}.

Much work has focused on the existence of scavenging in dinosaurs by using simple energetics approaches that typically focused on a single species namely \textit{Tyrannosaurus rex} \citep{ruxton2003could,carbone2011intra} but a recent modelling study investigated the likely prevelance of scavenging across a range of body sizes.
In it the authors demonstrated that species of intermediate body masses (approximatively 500 kg) would have gained the most benefit from scavenging.
This was the result of gut capacity limitations and the effects of competition at the carcass.
At the larger extreme this owes to the fact that gut capacity doesn't scale isometrically with body mass so the benefits of greater mass level off; there's only so much food an individual can consume at a single sitting \citep{calder1996size}.
For the smaller species, larger competitors would have prevented their access to carrion.

In Mesozoic systems some extremely large theropod dinosaurs had a morphology which suggests an ability to process bone e.g. the robust skull and dentition of \textit{T. rex} \citep{hone2010feeding}.
There is direct evidence that \textit{T. rex} did this in the form of distinctive wear marks on its tooth apices \citep{farlow1994wear,schubert2005wear} and the presence of bone fragments in its coprolites \citep{chin1998king}.
The animal also had an enormous bite force, with one estimate putting it at 57000 Newtons \citep{bates2012estimating}.
This is noted as being powerful enough to break open skeletal material \citep{rayfield2001cranial}.

\subsection*{Environment}
A major difficulty for terrestrial scavengers is competition with vultures.
Noctural behaviour in the Hyaenidae in general has been put forth as an adaptation to reduce competition with these exclusively diurnal birds \citep{gittleman2013carnivore}.

Osteophagy may have been even more viable during the Mesozoic era because the body mass distribution of herbviores tended to be skewed towards larger sizes \citep{10.1371/journal.pone.0051925}.
When we couple this with the fact that skeletal mass scales greater than linearly with body mass \citep{prange1979scaling} there would have been a lot of bone material to consume in the environment provided an animal had the biology to process it \citep{chure1997one}.
As we discussed earlier, osteophagy could be extremely beneficial to a scavenger.

The absence of flying vertebrates in the Palaeozoic may have permitted terrestrial forms to take in a higher proportion of carrion in their diet.
As noted earlier, a system similar to the Serengeti in productivity could have supported a terrestrial scavenger \citep{ruxton2004obligate}.
Indeed, in systems that were dominated by large ectothermic or mesothermic vertebrates 
the same primary productivity would have supported a greater biomass \citep{mcnab2009resources}.
The upshot of this is there was a higher biomass of herbivores dying and offering scavenging opportunities.
Predators were large bodied too compared to extant mammalian predators \citep{mcnab2009resources}, and so, especially if they were ectothermic, could last longer between meals rendering scavenging a more attractive behaviour relative to predation.

%TG: To be modified and inserted somewhere:
Scavenging behaviour might have evolved as soon as the first terrestrial tetrapods evolved.
In fact, some of the earlier tetrapods tracks dating back to the early Middle Devonian (393.3 - 387.7 Mya) were found in intertidal environments \citep{Niedzwiedzki2009}.
These environments are isolated from marine systems twice a day leaving potential carrion unexploited by marine vertebrates.
\cite{Niedzwiedzki2009} suggests that these environments ``would thus have allowed marine ancestors of tetrapods gradually to acquire terrestrial competence while accessing a new and essentially untouched resource.''
%TG: orginal sentence "The intertidal environment provides a ready food source of stranded marine animals on a twice-daily basis, in the immediate vicinity of the sea, and would thus have allowed marine ancestors of tetrapods gradually to acquire terrestrial competence while accessing a new and essentially untouched resource."

% and sensory perception \citep{farlow1994speculations}.
